# IPM Construction {#IPMCon}



##	Open population model with 2 age classes

### Model description
Population dynamics are represented using a female-based age-structured open 
population model with a pre-breeding census in spring. At census, females are 
divided into two age classes: "yearlings" (1-year old birds hatched during the 
breeding season of the previous year) and "adults" (birds older than one year). 
The motivation underlying this distinction is that reproductive output often 
differs for these two age classes in passerine birds. 
The dynamics of the female segment of the population over the time-interval from
census in year $t$ to census in year $t+1$ can be described with classic matrix
notation [@Caswell2001] as:

\
$$N_{tot,t+1} = \begin{bmatrix} N_{Y,t+1} \\ N_{A,t+1} \end{bmatrix} =
  \begin{bmatrix}
0.5F_{Y,t}sJ_t & 0.5F_{A,t}sJ_t \\
sA_t & sA_t
\end{bmatrix}\begin{bmatrix} N_{Y,t} \\ N_{A,t} \end{bmatrix} +
  \begin{bmatrix} Imm_{Y,t+1} \\ Imm_{A,t+1} \end{bmatrix}$$
\

$N_{tot,t+1}$ represents the total number of yearling and adult females in the 
population upon arrival in the breeding areas in year t. The total female 
population size, $N_{tot,t+1}$, is the sum of the numbers of yearling and adult 
females in the population in year $t+1$ ($N_{Y,t+1}$ and $N_{A,t+1}$, 
respectively) and consists of local survivors and recruits from the previous 
breeding season, as well as immigrant yearling ($Imm_{Y,t+1}$) and adult 
($Imm_{A,t+1}$) females. \

$F_{a,t}$ represents the expected number of fledglings produced by age class $a$
females during the breeding season in year $t$ and is the product of several 
vital rates. First, females in age class a may breed in a nestbox with 
probability $pB_{a,t}$ upon arrival to breeding areas in year $t$. Each breeding 
female may then lay a clutch containing a certain number of eggs (expected 
number = $CS_{a,t}$), and each egg within the clutch may hatch and survive to 
fledging. The probability of an egg hatching and surviving to fledging is 
divided into an age-independent probability of nest success ($pNS_t$, 
probability of complete clutch failure = $1-pNS_t$) and a survival probability 
of every egg/chick to fledging provided that the nest has not failed entirely 
($sN_{a,t}$, with $a$ = age of the mother). Consequently, the expected number of
fledglings produced by age class $a$ females in year t is defined as: 

\begin{equation}
F_{a,t}= pB_{a,t}\times CS_{a,t}\times pNS_{t}\times sN_{a,t}
\end{equation}

Fledglings that survive to the next breeding season and remain within the 
population (probability = $sJ_t$) contribute to next year’s yearling class 
($N_{Y,t+1}$). Yearlings and adults that survive to the next breeding season and
remain within the population (probability = $sA_t$) become part of next year’s 
adult age class ($N_{A,t+1}$).

### Code implementation including demographic stochasticity
Population process models within IPMs are typically implemented as stochastic
models that account for randomness in the outcomes of demographic processes at
the individual level ["demographic stochasticity", @Caswell2001; @kery2011].
The model described here is no different, meaning that the numbers of breeders, 
fledglings, and survivors are treated as binomial and Poisson random variables. 

Reproduction is modelled via two sets of random variables: a binomial random
variable representing the number of breeders in age class $a$ in year $t$, 
$B_{a,t}$ and a Poisson random variable representing the 
number of fledlings produced by breeders of age class $a$ in year $t$, 
$Juv_{a,t}$ . The implementation in the BUGS language used in the 
SPI-IPM code ([`IPMSetup.R`](https://github.com/SPI-Birds/SPI-IPM/blob/main/SPI-IPM_Code/02-04_IPM_Setup&Run/IPMSetup.R), lines 231-241) looks like: 
```{r eval=FALSE}
for (t in 1:Tmax){
  for(a in 1:A){
    
    ## 1) Breeding decision
    B[a,t] ~ dbin(pB[a,t], N[a,t])
    
    ## 2) Offspring production
    Juv[a,t] ~ dpois(B[a,t]*CS[a,t]*pNS[t]*sN[a,t]*0.5)
  }
}
```
In the code, the age indeces $a=1$ and $a=A=2$ correspond to yearlings and 
adults, respectively. 

Analogous to breeders, the numbers of local survivors -- both fledglings surviving their
first year and becoming yearlings, and yearlings and adults surviving to the 
next year -- are implemented as binomial random variables ([`IPMSetup.R`](https://github.com/SPI-Birds/SPI-IPM/blob/main/SPI-IPM_Code/02-04_IPM_Setup&Run/IPMSetup.R), lines 243-255):
```{r eval=FALSE}
for (t in 1:(Tmax-1)){
  
  ## 3) Annual survival of local birds
  # Juveniles -> Yearlings
  localN[1,t+1] ~ dbin(sJ[t], sum(Juv[1:A,t]))
  # Yearlings/Adults -> adults
  localN[2,t+1] ~ dbin(sA[t], sum(N[1:A,t]))
  
  ## 4) Immigration
  for(a in 1:A){
    N[a,t+1] <- localN[a,t+1] + Imm[a,t+1]
  }
}
``` 	  
Immigrant numbers are also treated as outcomes of stochastic processes, and 
these are detailed in [2.2.5 Immigrant count data likelihood](###	Immigrant count data likelihood).
  
##	Data likelihoods
IPMs obtain information on the population model's parameters (population sizes 
and vital rates) from several different data sets. Information in each data set
is channeled into model parameters via one or multiple data likelihoods. \
The SPI-IPM contains five data modules consisting of a total of eight data
likelihoods: nest count data (one likelihood), clutch size data (two likelihoods),
fledgling count data (three likelihoods), mark-recapture data (one likelihood), 
and immigrant count data (one likelihood). 
The likelihoods contained in each data module are described in detail in the 
following sub-chapters. The underlying data sets are introduced in 
[Chapter 1](#DataPrep) of the manual.

###	Nest count data likelihood
The population model defines the true size of the female segment of the breeding 
population in any year $t$ via the year- and age-specific number of breeding 
females:

\begin{equation}
B_{a,t}  \sim Binomial(N_{a,t}, pB_{a,t})
\end{equation}

The total size of the female breeding population (the sum of $B_{a,t}$  over all 
age classes) in year $t$ is expected to correspond closely with the observed 
number of first clutches laid in any year ($NestCount_t$). 
The breeding population count thus contains information about both breeding 
probability ($pB_{a,t}$) and population size ($N_{a,t}$) and its likelihood can 
be defined as:

\begin{equation}
NestCount_t  \sim Poisson((B_{Y,t} + B_{A,t}) \times NS\_Data_t)
\end{equation}

Here, the year-specific variable $NS\_Data_t$ is a correction factor introduced
for dealing with years in which no nest survey data was collected 
($NestCount_t = 0$ due to lack of sampling). $NS\_Data_t$ is set to 0 for years 
without data collection, and to 1 in all other years. 

The likelihood is written almost identically in the code, the only difference
being the use the `sum()` function over age classes $1$ to $A=2$ when 
specifying the Poisson distribution (this provides more flexibility for 
extending the model to more than two age classes): 
```{r eval=FALSE}
for(t in 1:Tmax){
  NestCount[t] ~ dpois(sum(B[1:A,t])*NS_Data[t])
}
``` 

###	Clutch size data likelihoods
The counting of incubated eggs provides information about both individual-level 
clutch size ($CS_{a,t}$) and reproductive output at the population level. 
Consequently, two separate likelihoods can be specified for within the clutch
size data module. 

Each individual clutch size observation can be treated as the outcome of a 
Poisson process with an expected value of $CS_{a,t}$ (where $a$ is the age of 
the female that laid the clutch, and $t$ the year in which the clutch was laid).
In the code, the likelihood is formulated for each clutch size observation $x$ 
(out of a total of $CS\_X$ observations) and includes nested indexing of the
expectation using data on female age (`CS_FAge`) and year (`CS_year`):
```{r eval=FALSE}
for(x in 1:CS_X){
	ClutchSize[x] ~ dpois(CS[CS_FAge[x], CS_year[x]])
}
``` 
Since both year and female age need to be part of the provided data, the 
individual-level clutch size likelihood can only be used with complete 
observations, i.e. clutches for which both number of eggs and age of the mother
are known (year should always be known). 

For the population-level likelihood, on the other hand, data on clutch sizes can
be included irrespective of whether the age of the mother is known. The total 
number of eggs counted in all nests laid in year $t$ can be described as:

\begin{equation}
EggNoTot_t \sim Poisson(sum(B_{Y:A,t} \times CS_{Y:A,t}) \times p_t^{EggNo})
\end{equation}

Since the product $B_{a,t}CS_{a,t}$ corresponds to all eggs laid in all 
nests/nestboxes within the study site, another correction factor ($p_t^{EggNo}$) 
is needed to account for the fact that eggs may not have been counted in all 
surveyed nests with breeding activity in each year. $p_t^{EggNo}$ 
(`EggNoSP[t]` in code) thus contains information of the year-specific proportion
of surveyed nests for which the numbers of eggs were counted. 

In code, the likelihood for the number of eggs at the population is written as:
```{r eval=FALSE}
for(t in 1:Tmax){

	# Expected "true" egg number (by mother age class)
	EggNo.ex[1:A,t] <- B[1:A,t]*CS[1:A,t]

	# Observed egg number (corrected by data availaility)
	EggNoTot[t] ~ dpois(sum(EggNo.ex[1:A,t])*EggNoSP[t])

}
```

###	Fledgling count data likelihoods
Analogous to observations of clutch size, counts of fledglings contain 
information on reproduction at both individual and population level.
Distributions of number of fledglings produced from a clutch are often 0-inflated
because incidents of harsh weather, predation, adandonment, etc. may result in 
complete loss of entire clutches. To account for this, we split data on fledgling
numbers and formulated separate likelihoods for the survival of the clutch as
a whole (probability of a nest not failing completely, $pNS_t$) and for each 
chick subsequently surviving to fledgling (probability $sN_{a,t}$). \
Whether or not a clutch succeeded (i.e. at least one chick survived to fledging,
`anyFledged`) was coded using 1 (success) and 0 (failure) and modelled as the 
outcome of a Bernoulli process with a year-dependent success probability $pNS_t$: 
```{r eval=FALSE}
for(x in 1:F_X){
  anyFledged[x] ~ dbern(pNS[F_year[x]])
}
```

For the subset of successful clutches, the number of fledglings produced
(`NoFledged`) was modelled as a binomial random variable with each egg laid in 
the clutch (`NoLaid`) having a probability of $sN_{a,t}$ (where $a$ = age of the
mother) to survive and fledge: 
```{r eval=FALSE}
for(x in 1:NoF_X){
	NoFledged[x] ~ dbin(sN[NoF_FAge[x], NoF_year[x]], NoLaid[x])
}
```

At the population level, both processes (nest success and survival to fledging
conditional on nest success) were combined into a single Poisson 
likelihood describing the total number of fledglings counted in the population
in a given year $t$ as the product of the number of breeding females, clutch
size, nest success, and survival to fledging: 

\begin{equation}
FledgedTot_t  \sim Poisson(sum(B_{Y:A,t}\times CS_{Y:A,t}\times pNS_t\times sN_{Y:A,t}) \times p_t^{Fledged})
\end{equation}

As in the clutch size data module ([Chapter 2.2.2](###	Clutch size data likelihoods)), 
a correction factor quantifying the proportion of nests for which data is 
availeble ($p_t^{Fledged}$, `FledgedSP[t]` in code) is used to account for 
missing records of fledgling numbers. 

In the code, calculation of the expected number of fledglings is based on the
expected number of eggs as calculated in the clutch size data module 
([Chapter 2.2.2](###	Clutch size data likelihoods)):
```{r eval=FALSE}
for(t in 1:Tmax){

	# Expected "true" fledgling number (by mother age class)
	Fledged.ex[1:A,t] <- EggNo.ex[1:A,t]*pNS[t]*sN[1:A,t]

	# Observed egg number (corrected by data availaility)
	FledgedTot[t] ~ dpois(sum(Fledged.ex[1:A,t])*FledgedSP[t])
}
```

###	Mark-recapture data likelihood

###	Immigrant count data likelihood



##	Priors and constraints

