# Modelling temporal variation {#TempVar}

Vital rates are expected to vary among years due to changes in environmental
conditions. `SPI-IPM` is set up to account for among-year variation in 
different (age-specific) vital rates $X_{a,t}$ using fixed effects of
supplied environmental covariates ($cov1_t$, $cov2_t$,...) and year random effects ($\epsilon_t^X$). The resulting generalized linear
mixed-models for time- and age-specific vital rates therefore take the following
form:
\begin{equation}
  link(X_{a,t}) = link(\mu_a^X) + \beta_{cov1}^X\times cov1_t + \beta_{cov2}^X\times cov2_t + ... + \epsilon_t^X
\end{equation}
Here, $\mu_a^X$ is the age-specific average vital rate (intercept) and $\beta_{cov1}$
and $\beta_{cov2}$ are the slopes for the effects of covariates $cov1$ and $cov2$
on the link scale, respectively. 
The link function depends on the vital rate,  and is set to logit for breeding 
probabilities ($pB_{a,t}$), nest success probabilities ($pNS_t$), and survival
probabilities ($sN_{a,t}$, $sJ_t$, $sA_t$) and log for clutch size ($CS_{a,t}$).


##	Random year variation
The random year effects included in the basic implementation of `SPI-IPM` are 
assumed to be normally distributed such that
\begin{equation}
  \epsilon_{t}^X \sim Normal(0, \sigma^X)
\end{equation}
where $\sigma^X$ is the standard deviation of random year effects on vital rate
$X_{a,t}$. 
The random effects in the basic implementation are also age-independent, meaning
that $\epsilon_t^X$ is included in the equations for both the yearling vital 
rate $X_{Y,t}$ and the adults vital rate $X_{A,t}$. The exception are the annual
survival rates for juveniles ($sJ_t$) and ($sA_t$), which both have separate
random effects because the drivers of survival variation are expected to vary 
between those two age classes, and the basic implementation includes an 
additional covariate effect $sJ_t$ only:
```{r eval=FALSE}
## Age- and time-dependent survival probabilities
logit(sJ[1:Tmax]) <- logit(Mu.sJ) + beta3.sJ*cov3[1:Tmax] + epsilon.sJ[1:Tmax]
logit(sA[1:Tmax]) <- logit(Mu.sA) + epsilon.sA[1:Tmax]

## Temporal random effects
for(t in 1:Tmax){
	 epsilon.sJ[t] ~ dnorm(0, sd = sigma.sJ)
	 epsilon.sA[t] ~ dnorm(0, sd = sigma.sA)
}

```
All random effects are treated as independent (= not correlated) in the basic 
implementation of the model, but the inclusion of correlation of random effects
across age-classes and/or vital rates is straightforward to implement^[Wheter or
not formally including random effects correlations is useful or not depends on
the biological questions of interest and the amoung of data available. When 
testing a model with correlated random effects for juvenile and adult survival 
on seven datasets from breeding populations of pied flycatchers in the UK, I
found that estimates did not differ from those obtained from a model with 
independent random effects, and the posterior distribution for the correlation
coefficient was so wide that no inference on strength or direction of the 
correlation was possible.] using e.g.
multivariate normal distributions or approaches similar to the one used in 
@nater2020.


##	Temporal covariates

The basic implementation of `SPI-IPM` features an example covariate model 
structure that was motivated by an analysis of populations of pied flycatcher
(\textit(Ficedula hypoleuca)) breeding in the UK. It involves three different
covariates ($cov1$, $cov2$, and $cov3$) that are assumed to affect nest success
probability ($pNS_{t}$), nestling survival ($sN_{a,t}$), and juvenile survival
($sJ_t$). However, it is very straightforward to alter the code to fit whatever 
alternative covariate structure is suitable for your particular analysis since
the inclusion different/additional continuous and categorical covariates always
works according to the same principles [see also @kery2011]. All covariates are
need to be passed to `SPI-IPM` as vectors or arrays. 

###	Continuous variables
Continuous variables are included as covariate effects using a specific slope 
parameter ($\beta$). Temporal effects then take the form 
$\beta_{cov1}\times cov1_t$, where $\beta_{cov1}$ is an estimated parameter 
quantifying strength and direction of the effect and $cov1_t$ is the value of
covariate $cov1$ at time $t$. 
In (generalized) linear mixed effects models as they are used in `SPI-IPM`, 
effects of several different covariates can just be added up on the link scale. 
For $pNS_t$, for example, this codes as^[Note that the implementation of this
in `IPMSetup.R` looks slightly different as it uses a vectorized formulation
(calculations done over all time-steps simultaneously instead of using a 
for-loop). Vectorized calculations are a nifty feature available in NIMBLE (but
not BUGS and JAGS); more on this in Chapter [4.1](##	Efficient implementation using NIMBLE).
]:
```{r eval=FALSE}
for(t in 1:Tmax){
  logit(pNS[t]) <- logit(Mu.pNS) + beta1.pNS*cov1[t] + beta2.pNS*cov2[t] + epsilon.pNS[t]
}
```
The basic implementation of `SPI-IPM` includes the following covariate models:
\begin{align*}
  & logit(pNS_t) = logit(\mu^{pNS}) + \beta_{cov1}^{pNS}\times cov1_t + \beta_{cov2}^{pNS} \times cov2_t + \epsilon_t^{pNS} \\
  & logit(sN_{a,t}) = logit(\mu_a^{pNS}) + \beta_{cov1}^{sN}\times cov1_t + \beta_{cov2}^{sN} \times cov2_t + \epsilon_t^{sN} \\
  & logit(sJ_t) = logit(\mu^{sJ}) + \beta_{cov3}^{sJ}\times cov3_t + \epsilon_t^{sJ}
\end{align*}
All covariates are continuous annual variables that have been standardized and
centered (mean = 0, sd = 1) prior to analysis.
$cov1$ and $cov2$ represent environmental conditions during the incubation and 
nestling period and hence influence nest success ($pNS_t$) and nestling survival
($sN_{a,t}$). In the case of the latter, covariates are further assumed to have
the same magnitude of effect on the nests of yearling and adult females (i.e.
the $\beta$ parameters are independent of age). $cov3$, on the other hand, 
symbolizes environmental conditions after fledging which impact juvenile annual
survival ($sJ_t$). No covariate effects are included for the other vital rates.

###	Categorical variables
The basic `SPI-IPM` does not include any categorical covariates, but since such
covariates may be relevant to a wide range of questions (e.g. some of the points
raised in Chapter [8.2](##	Adapting the population model for your species/population)),
I briefly illustrate how they could be included into vital rate models. 

Generally, there are two approaches to modelling categorical covariates in this
context.

The first approach works analogous to the approach for continuous covariates,
i.e. it uses the form $\beta_{cov}\times cov_t$. This is most relevant for
binary categorical covariates that symbolize some sort of "on-off" process. An
example of this would be if you would like to model the effect of an experimental
treatment that has been performed in some years ($cov_t = 1$) but not others
($cov_t = 0$). Your binary covariate then works as a "switch" that determines 
whether or not the effect of the experimental treatment ($\beta_{cov}$) applied
in a given year $t$ or not since $link(X_t) = link(\mu^X) + \beta_{cov_t}\times 
cov_t$ becomes $link(\mu^X) + \beta_{cov_t}$ when $cov_t = 1$ and $link(\mu^X)$
when $cov_t = 0$.

The second approach works via (nested) indexing and is more flexible since it
can technically account for any number of levels in your categorical covariate.
This can be relevant, for example, for categories of years ("good", "average",
"bad"), habitat types ("deciduous forest", "coniferous forest"), or individuals 
("male", "female"). 
The approach still uses $\beta$ parameters, but instead of multiplying the 
$\beta$ with the covariate value, we index the $\beta$ by the covariate value
such that $\beta_1$ corresponds to the effect of category 1 ($cat = 1$), 
$\beta_2$ corresponds to the effect of category 2 ($cat = 2$), and so on:

\begin{equation}
  link(X_{cat}) = link(\mu^X) + \beta_{cat}
\end{equation}
Priors then need to be provided for each category-specific $\beta$.

In practice, `SPI-IPM` still requires vital rates $X$ to be indexed by age class
and year (at least with the population model described in Chapter [2](# IPM Construction)).
That's where nested indexing becomes relevant. 
The relationship of a vital rate $X_{a,t}$ with a categorical year covariate can, 
for example, be coded as follows:
```{r eval=FALSE}
for(a in 1:A){
  for(t in 1:Tmax){
      log(X[a,t]) <- log(Mu.X[a]) + beta[cov[t]]
  }
  Mu.X ~ dunif(0, 10)
  beta ~ dunif(-5, 5)
}
```
where `cov[t]` is a vector of integer numbers that represent the different year
categories. 

Introducing categorical effects that rely on additional structure beyond year 
and age (for example effects of sex or location) requires changing the underlying
population model. Such extensions are currently not implemented in `SPI-IPM`, 
but see Chapter [8](# Useful extensions and outlook) for some perspectives.

###	Imputation of missing covariate values
Perhaps you have been wondering about how to deal with NAs in your covariate 
data? The good news is that `SPI-IPM` (just like any other Bayesian 
hierarchical model) can accommodate NAs in both continuous and categorical 
covariates. The (perhaps) less good news is that how well it works really 
depends on how large a proportion of your covariate data is NA. 

There are three practical requirements for working with partially observed 
covariate data:

1. Your covariate data containing numbers for your observed covariate values and NAs for your unobserved/unknown covariate values

2. Initial values with the same dimensions as your covariate data containing
numbers in the positions of NA covariate values and NAs in the positions of 
observed covariate values.
  

3. A model describing the distribution of missing covariate values. 

Numbers 1. and 2. are pretty self-explanatory (but see Chapter [4.2](##	Simulation of initial values) for more information on sampling initial values). 
Number 3. is going to depend on what type of covariate data you are dealing with. 

The basic implementation of `SPI-IPM` is set up to be able to deal with NA 
values in the continuous temporal covariates $cov1$, $cov2$, and $cov3$. These
covariates are assumed to have been standardized and centered, i.e. they should
more or less follow a $Normal(mean= 0, sd = 1)$ distribution. If we assume that
the observed and unobserved covariate values follow the same distribution (i.e.
the missing values are a random subset of all values), this can be used to 
specify the process model for the missing covariates in the code: 
```{r eval=FALSE}
for(t in 1:Tmax){
  cov1[t] ~ dnorm(0, sd = 1)
  cov2[t] ~ dnorm(0, sd = 1)
  cov3[t] ~ dnorm(0, sd = 1)
}
```
There are numerous alternatives for specifying distributions of missing values
in continous covariates, and they can be accommodated by changing the above 
section in the code. 

The number of candidate distributions are a bit more limited when there are 
missing values in categorical covariates. 
Chapter [8.2.1](###	Including partially observed age information) outlines an 
example for dealing with partially missing information on individual age. It
may also be helpful to remember that models with partially observed categorical
variables are essentially "mixture models" including auxiliary data about the
underlying distribution. 

##	Notes on covariate selection 
[TBA later]


